---
title: "Generate Fusion Summary Files"
output: html_notebook
author: Daniel Miller (D3b), Jaclyn Taroni (CCDL), Jo Lynne Rokita (D3b)
date: 2020, 2023
params:
  ci_run:
    label: "1/0 to run in CI"
    value: 1
    input: integer
editor_options: 
  chunk_output_type: inline
---

Generate fusion files specifically for consumption by molecular subtyping analyses

## Set up

### Libraries and functions

```{r}
library(tidyverse)
library(annoFuseData)
```

```{r}
#' Generate filtered fusion frame
#' @param df Unfiltered fusion data frame
#' @param bioid List of biospecimen IDs
#' @param fuses List of explicit fusion names
#' @param genes List of gene names
#' @return the filtered fusion data frame
filterFusion <- function(df, bioid, fuses, genes) {
  if (!missing(bioid)) {
    df <- dplyr::filter(df, Sample %in% bioid)
  }
  if (!missing(fuses) & !missing(genes)) {
    df <- dplyr::filter(df, FusionName %in% fuses |
                   Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  } else if (!missing(fuses)) {
    df <- dplyr::filter(df, FusionName %in% fuses)
  } else if (!missing(genes)) {
    df <- dplyr::filter(df,
                 Gene1A %in% genes |
                   Gene2A %in% genes |
                   Gene1B %in% genes |
                   Gene2B %in% genes)
  }
  return(df %>% 
           dplyr::select(Sample, FusionName, Fusion_Type, Gene1A, Gene1B, Gene2A, Gene2B, Gene1A_anno, Gene1B_anno, reciprocal_exists) %>%
           distinct())
}

#' Generate matrix with fusion counts
#' @param fuseDF Filtered fusion data frame
#' @param bioid List of biospecimen IDs that should be included in final table
#' @return Data frame that contains fusion counts
prepareOutput <- function(fuseDF, bioid) {
  fuseDF %>% 
    # some fusions have in-frame and frameshift fusion calls for a sample
    # this will make unique fusionName and Sample dataset to get 1/0 values
    dplyr::select(Sample,FusionName) %>%
    unique() %>%
    reshape2::dcast(Sample ~ FusionName,fun.aggregate = length) %>%
    right_join(data.frame(Sample = bioid)) %>%
    replace(is.na(.), 0) %>%
    rename(Kids_First_Biospecimen_ID = Sample)
}
```

### Read in data

```{r}
dataDir <- file.path("..", "..", "data")
fusDir <- file.path("..", "..", "analyses", "fusion_filtering", "results")
annotDir <- file.path("..", "..", "analyses", "fusion_filtering", "references")
inputDir <- "input"
#' Annotation file to be used for identifying kinase genes
annot <- read.delim(system.file("extdata", "genelistreference.txt", package = "annoFuseData"))

#' The putative oncogenic fusion file is what we'll use to check for the #' presence or absence of the fusions.
#' However, some biospecimens are not represented in this filtered, prioritized
#' file but have been assayed - we will gather those from the histology file
hist <- read_tsv(file.path(dataDir, "Hope-GBM-histologies-base.tsv"))

# read fusion data

putativeOncogenicDF <- 
  read_rds(file.path(dataDir, "Hope-fusion-putative-oncogenic.rds")) %>% 
  rename(Sample = Kids_First_Biospecimen_ID)


```


### Output

```{r}
lgg_hggFile <- file.path(inputDir, "fusion_summary_hgg_foi.tsv")
```

## Fusions and genes of interest

Taken from [`AlexsLemonade/OpenPBTA-analysis#245`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/245), [`AlexsLemonade/OpenPBTA-analysis#251`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/251), [`AlexsLemonade/OpenPBTA-analysis#623`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/623) respectively, and [`AlexsLemonade/OpenPBTA-analysis#808`](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/808) 

```{r}
#' Exact match a list of fusions common in low-grade glioma and high-grade glioma tumors
#' as well as fusions containing a particular gene with any other gene
lggFuses <- c(
  "KIAA1549--BRAF",
  "FGFR1--TACC1",
  "MYB--QKI"
)
lgghggGenes <- c(
  "BRAF",
  "ALK",
  "ROS1",
  "NTRK1",
  "NTRK2",
  "NTRK3",
  "PDGFRA",
  "FGFR2",
  "FGFR1",
  "MYB",
  "MYBL1",
  "RAF1", 
  "MET"
)

```

### Filter putative oncogenic fusions list

```{r}

allFuseLGGHGG <- filterFusion(df = putativeOncogenicDF,
                             fuses = lggFuses,
                             genes = lgghggGenes)

```

Get the biospecimen IDs that have had either RNA-Seq or Targeted fusion panel performed.
We use the putative oncogenic file here, therefore any sample that was profiled but does not have a fusion that is relevant to the subtyping tickets is not _missing_ but instead has no evidence of the relevant fusions.

```{r}
# gather and combine
specimensUnion <- hist %>%
  filter(experimental_strategy == "RNA-Seq") %>%
  pull(Kids_First_Biospecimen_ID)
```


#### Perform selection for LGG and HGG fusions
First pull the fusions or genes from the goi list which are not kinases for the final output file, since these will not need further interrogation.
```{r}
# Which genes/fusions are not kinases, but in the list?
# Separate fusions into genes, combine with gene list, check for not kinase
lggFuses_df <- as.data.frame(lggFuses) %>%
  separate(lggFuses, into = c("Gene1A", "Gene1B"), remove = F)

kinases <- annot %>%
  filter(type == "Kinase") %>%
  pull(Gene_Symbol)

lggFuses_df <- lggFuses_df %>% 
  dplyr::mutate(Gene1A_anno = case_when(Gene1A %in% kinases ~ "Kinase", TRUE ~ "Non-Kinase"), 
                Gene1B_anno = case_when(Gene1B %in% kinases ~ "Kinase", TRUE ~ "Non-Kinase"))


```

## as we do not have the columns to show whether the domain is retained, so I deleted the 3' and 5' part, and direct use 

#### Write LGG and HGG fusions to file

```{r}
allFuseLGGHGG %>%
  rename(Kids_First_Biospecimen_ID = Sample) %>%
  arrange(Kids_First_Biospecimen_ID) %>%
  write_tsv(lgg_hggFile)
```

## Session Info

```{r}
sessionInfo()
```
