---
title: "High-Grade Glioma Molecular Subtyping - Defining Lesions"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Chante Bethell for ALSF CCDL
date: 2019
---

This notebook looks at the defining lesions for all samples for the issue of 
molecular subtyping high-grade glioma samples in the OpenPBTA dataset. 

# Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/molecular-subtyping-HGG/01-HGG-molecular-subtyping-defining-lesions.Rmd', clean = TRUE)"
```

# Set Up

```{r}
library(tidyverse)

# Get `magrittr` pipe
`%>%` <- dplyr::`%>%`
```

## Directories and Files

```{r}
# Detect the ".git" folder -- this will in the project root directory.
# Use this as the root directory to ensure proper sourcing of functions no
# matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# Get hgg subset folder from the module in master
hgg_subset_dir <- file.path(root_dir, 
                        "analyses",
                        "molecular-subtyping-HGG",
                        "hgg-subset")

# File path to results directory
results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-HGG", "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Read in metadata
metadata <-
  readr::read_tsv(file.path(root_dir, "data", "HOPE-GBM-histologies-base.tsv"), 
                  guess_max = 100000) %>%
  dplyr::filter(!is.na(pathology_diagnosis))

# Read in the JSON file that contains the strings we'll use to include or
# exclude samples for subtyping - see 00-HGG-select-pathology-dx
path_dx_list <- jsonlite::fromJSON(
  file.path(hgg_subset_dir,
            "hgg_subtyping_path_dx_strings.json")
)

#### Filter metadata -----------------------------------------------------------

# Filter metadata based on pathology diagnosis fields and include samples that
# should be classified as high-grade glioma based on defining lesions

# Samples included on the basis of the pathology diagnosis fields
hgg_only_meta <- metadata %>%
  # Inclusion on the basis of CBTTC harmonized pathology diagnoses
  filter(pathology_diagnosis %in% path_dx_list$exact_path_dx |
         # Inclusion based on pathology free text diagnosis
         pathology_free_text_diagnosis %in% path_dx_list$path_free_text_exact |
         # Inclusion based on pathology free text diagnosis for IHG
         pathology_free_text_diagnosis %in% path_dx_list$IHG_path_free_path_dx)

dna_bsids <- hgg_only_meta %>%
  filter(!is.na(pathology_diagnosis),
                (experimental_strategy %in% c("WGS", "WXS", "Targeted Sequencing") & is.na(RNA_library))) %>%
  pull(Kids_First_Biospecimen_ID)

```

# Prepare Data
# Read in snv consensus and hotspots mutation data, filtering out LGAT
 - We will use the `snv-consensus-plus-hotspots.maf.tsv.gz` which contains consensus SNV calls that are present in all 3 callers (strelka2, mutect2, and lancet) and with hotspot rescued. Also include the DGD MAF, `snv-dgd.maf.tsv.gz` for OpenPedCan.  
 - Calls that overlap MSKCC hotspots found in any caller are included (except if only vardict calls the site as variant, we remove these calls since we have a lot of calls unique to vardict which we consider as false positive as discussed [here](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/snv-callers#snv-caller-comparison-analysis))

```{r}

# select tumor sample barcode, gene, short protein annotation and variant classification
keep_cols <- c("Chromosome",
             "Start_Position",
             "End_Position",
             "Strand",
             "Variant_Classification",
             "IMPACT",
             "Kids_First_Biospecimen_ID",
             "Hugo_Symbol",
             "HGVSp_Short",
             "Exon_Number")
# DGD MAF
snv_consensus_hotspot_maf <- readr::read_rds(
  file.path(root_dir, "data" , "Hope-consensus-mutation.maf.rds")) %>%
  select(all_of(keep_cols))
                                  

```


## SNV consensus mutation data - defining lesions

```{r}
# Filter the snv consensus mutation data for the target lesions
snv_lesions_df <- snv_consensus_hotspot_maf  %>%
  dplyr::filter(Hugo_Symbol %in% c("H3-3A", "H3C2",
                                   "H3C3", "H3C14") &
                  HGVSp_Short %in% c("p.K28M", "p.G35R",
                                     "p.G35V")) %>%
  dplyr::select(Kids_First_Biospecimen_ID, Hugo_Symbol, HGVSp_Short) %>%
  dplyr::mutate(
    `H3F3A_H3-3A.K28M` = dplyr::case_when(Hugo_Symbol == "H3-3A" &
                                    HGVSp_Short == "p.K28M" ~ "Yes",
                                  TRUE ~ "No"),
    HIST1H3B_H3C2.K28M = dplyr::case_when(
      Hugo_Symbol == "H3C2" & HGVSp_Short == "p.K28M" ~ "Yes",
      TRUE ~ "No"),
    HIST1H3C_H3C3.K28M = dplyr::case_when(
      Hugo_Symbol == "H3C3" & HGVSp_Short == "p.K28M" ~ "Yes",
      TRUE ~ "No"),
    HIST2H3C_H3C14.K28M = dplyr::case_when(
      Hugo_Symbol == "H3C14" & HGVSp_Short == "p.K28M" ~ "Yes",
      TRUE ~ "No"),
    `H3F3A_H3-3A.G35R` = dplyr::case_when(Hugo_Symbol == "H3-3A" &
                                    HGVSp_Short == "p.G35R" ~ "Yes",
                                  TRUE ~ "No"),
    `H3F3A_H3-3A.G35V` = dplyr::case_when(Hugo_Symbol == "H3-3A" &
                                    HGVSp_Short == "p.G35V" ~ "Yes",
                                  TRUE ~ "No")
  ) %>%
  dplyr::select(
    -HGVSp_Short,
    -Hugo_Symbol
  )

# add back in samples with no evidence of these specific mutations and are not LGAT
snv_lesions_df <- snv_lesions_df %>%
  dplyr::bind_rows(
    data.frame(
      Kids_First_Biospecimen_ID = setdiff(dna_bsids, snv_lesions_df$Kids_First_Biospecimen_ID)
    )
  ) %>%
  dplyr::mutate_all(function(x) tidyr::replace_na(x, "No"))
```

Add a column that keeps track of the presence of any defining lesion.
We'll use this to create subset files in the next step.

```{r}
snv_lesions_df <- snv_lesions_df %>%
  dplyr::mutate(
    defining_lesion = dplyr::case_when(
      `H3F3A_H3-3A.K28M` == "Yes" ~ TRUE,
      HIST1H3B_H3C2.K28M == "Yes" ~ TRUE,
      HIST1H3C_H3C3.K28M == "Yes" ~ TRUE,
      HIST2H3C_H3C14.K28M == "Yes" ~ TRUE,
      `H3F3A_H3-3A.G35R` == "Yes" ~ TRUE,
      `H3F3A_H3-3A.G35V` == "Yes" ~ TRUE,
      TRUE ~ FALSE
    )
  )
```

## What are the diagnoses of the samples which have defining lesions but are not HGG, LGG, or EPN? Make sure to add these back at the end!

```{r non HGG samples with defining lesions}
metadata %>%
  filter(Kids_First_Biospecimen_ID %in% setdiff(snv_lesions_df$Kids_First_Biospecimen_ID, dna_bsids)) %>%
  select(Kids_First_Biospecimen_ID, sample_id, pathology_diagnosis)

```


Add other identifiers and sort.

```{r}
snv_lesions_df <- metadata %>%
  dplyr::select(Kids_First_Participant_ID, 
                sample_id,
                Kids_First_Biospecimen_ID) %>%
  dplyr::inner_join(snv_lesions_df,
                    by = c("Kids_First_Biospecimen_ID" = "Kids_First_Biospecimen_ID")) %>%
  dplyr::arrange(Kids_First_Participant_ID, sample_id)
```



## Save final table of results

```{r}
# Save final data.frame to file
readr::write_tsv(snv_lesions_df,
                 file.path(results_dir, "Hope_HGG_defining_lesions.tsv"))
```

## Session Info

```{r}
# Print the session information
sessionInfo()
```
