---
title: "1p/19q co-deleted oligodendrogliomas"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Jaclyn Taroni for ALSF CCDL
date: 2020
---

This notebook will look at 1p/19q codeletions in the entire OpenPBTA cohort.

The purpose is to identify samples that should be classified as 1p/19q co-deleted oligodendrogliomas.

## Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/molecular-subtyping-HGG/08-1p19q-codeleted-oligodendrogliomas.Rmd', clean = TRUE)"
```

## Set up

### Libraries and functions

```{r}
library(tidyverse)
```

### Directories

```{r}
# File path to results directory -- this contains the cleaned data
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
results_dir <- file.path("results")
data_dir <- file.path("..", "..", "data")
scratch_dir <- file.path("..", "..", "scratch")
```

### Read in data

```{r}
# TODO: update once the consensus GISTIC results are in the data release
unzip(file.path(root_dir, "data", "cnv-consensus-gistic.zip"),
      exdir = file.path(root_dir, "data"),
      files = file.path("cnv-consensus-gistic", "broad_values_by_arm.txt"))

gistic_df <- data.table::fread(file.path(root_dir,
                                         "data",
                                         "cnv-consensus-gistic",
                                         "broad_values_by_arm.txt"),
                               data.table = FALSE)
```

## Look at features

### 1p/19q codeletions

```{r}
relevant_arms_df <- gistic_df %>%
  filter(`Chromosome Arm` %in% c("1p", "19q")) %>%
  column_to_rownames("Chromosome Arm") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Kids_First_Biospecimen_ID")
```

```{r}
codeleted_samples <- relevant_arms_df %>%
  filter(`1p` < 0, `19q` < 0)

codeleted_samples
```

Using the consensus GISTIC results, we have no instances of 1p, 19q co-deletion.

## Session Info

```{r}
sessionInfo()
```

